Bee brain size is positively associated with urban lifestyles but not with positive population trends

### 2022-11-25 Manuscript/Ms.Rmd
Update on methods

## (a) Overview

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE}

@@

## (b) Bee species dataset

@@

## (c) Brain and body size measurements

@@

## (d) Occurrence data

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE}

@@

We downloaded occurrence information for all species from the Global Biodiversity Information Facility (GBIF; http://www.gbif.org/) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. Because the records were biased towards some areas that were associated with political boundaries, we selected the states or countries with highest density of records for our set of species. For North America, we selected states located on the East Coast of United States (Figure 1A) that covered an approximate area of `r usakm2` km$^{2}$. These states were: Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. Similarly, we selected countries located on the north and center of Europe (Figure 1B) that represented a total area of `r eukm2` km$^{2}$. These countries were: England, Belgium, Netherlands, Germany and Denmark. All occurrence data used in this study can be downloaded from https://doi.org/10.15468/dl.nenuem where just the countries of interest were retrieved. To avoid biases in our analysis, we filtered the occurrence records according the following criteria: (i) we only included species with a widespread distribution of occurrences along our study area; (ii) all species had a minimum number of 50 records; (iii) all duplicated records were excluded (i.e., records of specimens with the same date and location); (iv) we only included georeferenced records with a minimum of two decimals of latitude/longitude coordinates; and (v), we only included records between the time interval 1990 and 2022 in order to avoid large  differences with the extracted land cover classifications for United States and Europe (2006).

## (e) Land cover classification

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center", fig.height=12, fig.width=12, fig.cap="\\textbf{Figure 1.} Maps showing the study areas and the number of GBIF occurrences in logarithmic scale for the selected bee species (N = 85). A) East Coast of United States with records for Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. B) European continent with records downloaded for England, Belgium, Netherlands, Germany and Denmark.", fig.pos="p"}

@@

@@

## (f) Construction of the phylogenetic tree

@@

The quantitative data of population trends for North American bees were extracted from Bartomeus et al., 2013, where they analyzed trends for 187 species over time, using data from pinned bees from different North American museums and universities. We assigned the estimated value of their relative change over time models as values of population trends for the 44 species for which we have brain size measurements (positive for increasing populations and negative for decreasing populations). Habitat preferences for USA bee data were extracted from Collado et al., 2019, were habitat preference for forests and urban areas was calculated comparing species occurrence in each habitat to what would be expected if species occur at random using null models. This preference index ranges from 0 (avoidance) to 1 (preference).In order to have a representative value of brain weight and body size per species, we used means of the different collected individuals from each species. Twenty-four individuals with brain weight values one standard deviation higher or lower than the average were removed (see sup mat analysis) because human errors in their measurement are likely with such delicate tissues. Specially, brain weights can give errors due to dehydration or insufficient removal of fat bodies and trachea around the brain. For eusocial species, we only used workers. We estimated relative brain size using residuals from a log-log correlation of brain weight against body size to remove body size allometric relationships. (see Wurm & Fisicaro, 2014) (LM estimate ± SE = 1.88 ± 0.06, p \< 0.001, R2= 0.83, n = 163 species). Residuals were calculated using our complete brain weights and body size database.

@@

>>>>

## Overview

@@

```{r Taxonomic information about the dataset, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE}

@@

## (a) Bee species dataset

@@

## (b) Brain and body size measurements

@@

## (c) Occurrence data

```{r Loading study site areas, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE}

@@

```{r Species and occurrences info, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE}

#Read occurrence records and extract relevant info
#Load libraries
library(data.table)
#Load clean data with bee occurrences
europe <- read.table("../Data/Europe_data/all_above_50_europe.csv.gz",  header=T, quote="\"", sep=",")
usa <- read.table("../Data/Usa_data/all_above_50_usa.csv.gz",  header=T, quote="\"", sep=",")

#Europe
eu_occurrences = prettyNum(nrow(europe), big.mark=",")
records_eu = europe %>% 
group_by(species) %>% 
summarize(no_rows= length(species))

#USA
usa_occurrences = prettyNum(nrow(usa), big.mark=",")
records_usa = usa %>% 
group_by(species) %>% 
summarize(no_rows= length(species))

#Number of species
usa_spp = nrow(records_usa)
eu_spp = nrow(records_eu)
spp_number = nrow(records_eu) + nrow(records_usa)
#Median number of records per dataset
eu_median = median(records_eu$no_rows)
usa_median = median(records_usa$no_rows)

We downloaded occurrence information for all species from the Global Biodiversity Information Facility (GBIF; http://www.gbif.org/) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. Because the records were biased towards some areas that were associated with political boundaries, we selected the states or countries with highest density of records for our set of species. For North America, we selected states located on the East Coast of United States (Figure 1A) that covered an approximate area of `r usakm2` km$^{2}$. These states were: Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. Similarly, we selected countries located on the north and center of Europe (Figure 1B) that represented a total area of `r eukm2` km$^{2}$. These countries were: England, Belgium, Netherlands, Germany and Denmark. All occurrence data used in this study can be downloaded from https://doi.org/10.15468/dl.nenuem where just the countries of interest were retrieved. To avoid biases in our analysis, we filtered the occurrence records according the following criteria: (i) we only included species with a widespread distribution of occurrences along our study area; (ii) all species had a minimum number of 50 records; (iii) all duplicated records were excluded (i.e., records of specimens with the same date and location); (iv) we only included georeferenced records with a minimum of two decimals of latitude/longitude coordinates; and (v), we only included records between the time interval 1990 and 2022 in order to avoid large differences with the extracted land cover classifications from 2006. After pruning the data, our final dataset comprised `r spp_number` species with  `r usa_occurrences` occurrence records for North America (`r usa_spp` species; median occurrences = `r usa_median`) and `r eu_occurrences` records for Europe (`r eu_spp` species; median occurrences = `r eu_median`).

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center", fig.height=12, fig.width=12, fig.cap="\\textbf{Figure 1.} Maps showing the study areas and the number of GBIF occurrences in logarithmic scale for the selected bee species (N = 85). A) East Coast of United States with records for Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. B) European continent with records downloaded for England, Belgium, Netherlands, Germany and Denmark.",  out.width="90%"}

@@

## (d) Land cover classification

We assigned a habitat type for all GBIF occurrences by merging the information from land cover inventories with the different georeferenced records. The land cover classification was obtained from the 2006 online inventories of the National Land Cover Database (NLCD) for United States and the Corine Land Cover (CLC) for Europe. After downloading these as raster files, we used the functions *rast* and *extract* from the *Terra* package version 1.5-21 to read and obtain the cover classification, respectively. Then, to simplify the interpretation and conduct a joint analysis for both occurrence datasets, we divided the resulting cover classes into three single categories: (i) natural, (ii) agricultural and (iii) urban. The original extracted cover classes and our classification into these three different categories are shown in Table S1 for United States and Table S2 for Europe.

## (e) Construction of the phylogenetic tree

## (f) Analysis

@@

@@

Habitat preferences for USA bee data were extracted from Collado et al., 2019, were habitat preference for forests and urban areas was calculated comparing species occurrence in each habitat to what would be expected if species occur at random using null models. This preference index ranges from 0 (avoidance) to 1 (preference).In order to have a representative value of brain weight and body size per species, we used means of the different collected individuals from each species. Twenty-four individuals with brain weight values one standard deviation higher or lower than the average were removed (see sup mat analysis) because human errors in their measurement are likely with such delicate tissues. Specially, brain weights can give errors due to dehydration or insufficient removal of fat bodies and trachea around the brain. For eusocial species, we only used workers. We estimated relative brain size using residuals from a log-log correlation of brain weight against body size to remove body size allometric relationships. (see Wurm & Fisicaro, 2014) (LM estimate ± SE = 1.88 ± 0.06, p \< 0.001, R2= 0.83, n = 163 species). Residuals were calculated using our complete brain weights and body size database.

@@

# Supplementary material

```{r Table S1, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center"}

#Create table of occurrences per habitat type (USA)
#Load libraries
library(tidyverse)
library(kableExtra)

#Load extracted data
pref <- read_csv("../Data/Usa_data/land_cover_usa.csv.gz") %>% 
dplyr::select(species, cover.names) %>% 
mutate_if(is.character,as.factor)

#Create table with original and new cover categories
pref %>% 
rename(Cover_names = cover.names) %>% 
select(Cover_names) %>% 
group_by(Cover_names) %>% 
summarise(Occurrences = length(Cover_names)) %>% 
mutate(Cover_names1 = fct_recode(as.factor(Cover_names),
       Discarded = "Barren Land", 
       Agricultural = "Cultivated Crops", 
       Natural = "Deciduous Forest", 
       Urban = "Developed, High Intensity", 
       Urban = "Developed, Medium Intensity",
       Urban = "Developed, Low Intensity",
       Urban = "Developed, Open Space", 
       Natural = "Emergent Herbaceous Wetlands", 
       Natural = "Evergreen Forest", 
       Agricultural = "Hay/Pasture", 
       Natural = "Herbaceous", 
       Natural = "Mixed Forest", 
       Discarded = "Open Water", 
       Natural = "Shrub/Scrub", 
       Natural = "Woody Wetlands")) %>% 
arrange(desc(Occurrences)) %>%  
kbl() %>% 
kable_styling(font_size = 12,latex_options = "hold_position")

```{r Table S2, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center"}

#Create table of occurrences per habitat type (USA)
#Load libraries
library(tidyverse)
library(kableExtra)

#Load extracted data
pref <- read_csv("../Data/Europe_data/land_cover_europe.csv.gz") %>% 
    dplyr::select(Species, Cover_names) %>% 
    mutate_if(is.character,as.factor)

#Create table with original and new cover categories
pref %>% 
select(Cover_names) %>% 
group_by(Cover_names) %>% 
summarise(Occurrences = length(Cover_names)) %>% 
mutate(Cover_names1 = fct_recode(as.factor(Cover_names),
      Urban = "Airports", #1
      Discarded = "Bare rocks", #2
      Natural = "Beaches, dunes, sands", #3
      Natural = "Broad-leaved forest", #4
      Natural = "Coastal lagoons", #5
      Agricultural = "Complex cultivation patterns", #6
      Natural = "Coniferous forest", #7
      Urban = "Construction sites", #8
      Urban = "Continuous urban fabric", #9
      Urban = "Discontinuous urban fabric", #10
      Discarded = "Dump sites", #11
      Discarded = "Estuaries", #12
      Agricultural = "Fruit trees and berry plantations", #13
      Urban = "Green urban areas", #14)
      Urban = "Industrial or commercial units", #15
      Natural = "Inland marshes", #16
      Natural = "Intertidal flats", #17
      Agricultural = "Land principally occupied by agriculture, with significant areas of natural vegetation", #18
      Discarded = "Mineral extraction sites", #19
      Natural = "Mixed forest", #20
      Natural = "Moors and heathland", #21
      Agricultural = "Natural grasslands", #22
      Agricultural = "Non-irrigated arable land", #23
      Agricultural = "Pastures", #24
      Natural = "Peat bogs", #25
      Urban = "Port areas", #26) 
      Discarded = "Road and rail networks and associated land", #27
      Natural = "Salt marshes", #28
      Discarded = "Sea and ocean", #29
      Natural = "Sparsely vegetated areas", #30
      Urban = "Sport and leisure facilities", #31
      Natural = "Transitional woodland-shrub", #32
      Agricultural = "Vineyards", #33
      Discarded = "Water bodies", #34
      Discarded = "Water courses" #35  
)) %>% 
mutate(Cover_names = fct_recode(as.factor(Cover_names),
      "Land principally occupied by agriculture" = "Land principally occupied by agriculture, with significant areas of natural vegetation")) %>% 
arrange(desc(Occurrences)) %>%  
kbl() %>% 
kable_styling(font_size = 12,latex_options = "hold_position")
